Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-05-25 pkg/paper/Paper/draft.Rnw
Rewrote modeling section, merging it with the GL appendix.
The intro to the results section has also been enhanced.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@100 b679b2f7-6539-45d0-a605-5ddcb976cde9

However, a deeper inquiry into this problem is in need of a method able to
capture changes in inflation dynamics. We contribute to this debate by
developing a new method for testing for structural breaks in the dynamics of the
inflation rates and thereby upon the effect of the EMU on a number of European
countries both within and outside the eurozone. The framework developed in this
paper makes it possible to test for any significant changes of a countries
inflation dynamics and thus helps interpreting the effects of the EMU. We focus
on changes in the first three moments of these inflation rates, which we assume
to be distributed according to a generalized logistic distribution (GL). The GL
distribution is a suitable one due to its flexibility to capture nonsymmetry and
because of its fatter tails. For this purpose, we develop a new estimation
approach building on previous research by \citet{z07}.

@@

Hence, a likelihood-based model is adopted that can also incorporate stylized
facts for return data, such as heavy tails and skewness. Here, the seasonally
adjusted returns $y_t$ ($t = 1, \dots, n$) are assumed to come from a
generalized logistic (GL) distribution \citep[see][]{johnson} with
three-dimensional parameter $\phi_t = (\theta_t, \sigma_t, \delta_t)^\top$ (for
location~$\theta_t$, scale~$\sigma_t$, and shape~$\delta_t$ at time $t$). Structural change
techniques are then employed to check whether the parameter vector $\phi_t$
remains constant over time and, if that is not the case, when and how the
parameters change. This framework allows to trace breaks in the
evolution of mean, variance and skewness of the inflation series which may
possibly be linked to underlying regime changes, e.g., in monetary policy.

@@

\fixme{Z: potentially say something about the subsequent subsections}

@@

The data suggested that the use of a normal distribution might cause some
estimation problems, since most of the time series exhibit asymmetric properties
(with sometimes rather strong skewness) and -- compared to the normal
distribution -- a fatter tail. To accommodate for this fact, the use of the
generalized logistic distribution (GL) with its relative low degree of
complexity, seemed appropriate. On the one hand, it is a simple distribution and
on the other hand, it is flexible enough to capture essential properties of the
data. For the inflation series $y_t = 100\cdot log(HICP_t/HICP_{t-1})$ we
therefore assume a GL distribution as defined in \citet{johnson}:

\begin{eqnarray}
f(y | \theta, \sigma, \delta) & = & \frac{\frac{\delta}{\sigma} \cdot \exp^{-\frac{y-\theta}{\sigma}}}{(1+\exp^{-\frac{y-\theta}{\sigma}})^{(\delta+1)}} 
\label{eq:dglogis}
\end{eqnarray}

with location $\theta$, scale $\sigma$ and shape $\delta$. For $\delta=1$ the
distribution simplifies to the logistic distribution, for $\delta<1$ it is
skewed to the left and for $\delta$$>$1 it is skewed to the right (see
Appendix~\ref{sec:gld2} for more details).

In econometrics, the logistic distribution is often used in income distributions
and growth models. This is due to its fatter tails, which fit these kind of
data somewhat better. \citet{won} develop a maximum likelihood \fixme{Jesus: is
this not better than what you do here: answer: no, they use completely other
framework} estimator if the underlying distribution is a GL distribution. A
similar GL distribution is also used in \citet{tolikas} who analyze extreme risk
and value-at-risk in the German stock market, although they do not use a shape
parameter. Regarding inflation rates, the GL distribution is -- to our best
knowledge -- only used in context with expected inflation.

\citet{batchelor} use a logistic distribution (not its generalization) to model
the distribution of mean expected inflation rates.

@@

Under the assumption that $y_t$ is distributed according to the generalized
logistic distribution, we test for the hypothesis that the parameter vector
$\hat{\phi}$ stays stable over time:

\begin{eqnarray}
& H_0: \phi_t = \phi_0 \quad (t=1,...,n) \nonumber 
\end{eqnarray}

against the $H_1$ of changes in at least one of the parameters. We first
estimate the parameters of the GL-distribution by means of maximum likelihood
$\hat{\phi} ~=~ \argmax_\phi \sum_{t=1}^n \log f(y_t|\phi)$. The parameter
stability is assessed using the empirical scores $s(y_t|\hat{\phi})$ as measures
of model deviation.\footnote{for details see Appendix~\ref{sec:gld2} and
\citet{z07}}  The empirical fluctuation process $\mathit{efp}(\cdot)$ captures
deviations from a hypothesized zero mean over time and is given by
%
\begin{eqnarray}
\mathit{efp}(t) ~=~ \hat{V}^{-1/2} n^{-1/2} \sum_{i=1}^{\lfloor nt \rfloor} s(y_i|\hat{\theta}, \hat{\sigma}, \hat{\delta}) \quad (0 \leq t \leq 1), \nonumber
\label{eq:efp}
\end{eqnarray}
%
where $\hat{V}$ is some consistent estimator of the variance of the scores,
which is used to decorrelate the series as to avoid problems due to serial
correlation. Then a functional central limit theorem (FCLT) for
$\mathit{efp}(\cdot)$ is employed, which converges to a 3-dimensional Brownian
bridge: $\mathit{efp}(\cdot) ~\overset{d}{\rightarrow}~ W^0(\cdot)$.

\subsection{Test}

After the $\mathit{efp}(\cdot)$ is computed, it conveys information about
departures from the $H_0$ of parameter stability over time. For a precise
quantification of the actual deviation from its hypothesized zero-mean path, the
main idea of the test statistics presented here is to aggregate the
$\mathit{efp}(\cdot)$ in such a way as to be able to construct a test statistics
so that critical values and p values can be derived. In this paper, we employ
the sup$\mathit{LM}$ test of \citet{andrews}. This test is well suited for
single break alternatives and performs better if more than just one of the
distribution parameters change.

It is given by:
%
\begin{eqnarray}
\sup_{t \in [0.1, 0.9]} \frac{\Vert \mathit{efp}(t) \Vert_2^2}{t(1-t)} \nonumber
\end{eqnarray}
%
where the appropriate $p$~values can be computed using from the corresponding
limiting distribution -- with $\mathit{efp}(t)$ replaced by $W^0(t)$.
The sup$\mathit{LM}$ test computes a test statistics for all possible change points in a
given time frame and rejects the null of no change if the maximum is too high.

@@

The dating of structural breaks builds upon the work of \cite{bp03}. If a break
is detected via the sup$\mathit{LM}$ test, the next step is estimating the number and the
timing of these break points. Breakpoints $\tau_1, \dots,\tau_B$ are estimated
via maximization of the full segmented likelihood
%
\begin{eqnarray}
\sum_{b=1}^{B+1} \sum_{t=\tau_b + 1}^{\tau_b} \log f(y_t | \phi^{(b)}) \nonumber
\end{eqnarray}
%
If the test statistics is significant (in this case the sup$\mathit{LM}$ test), at least
one break is estimated. In principle, the optimal segmentation (i.e. the
concrete dates of a break) can be computed once B (the number of breakpoints) is
known. However, B needs to be based on the observed data as well. The solution
is to compute the timing of the breaks for a number of different values for B
and then choose B by means of an information criterion, in this case the LWZ
(Liu, Wu and Zidek) as developed in \citet{liu}. All parameters $\tau_1, \dots,
\tau_B, \phi^{(1)}, \dots, \phi^{(B+1)}$ can be estimated jointly using dynamic
programming.\footnote{More details for this estimation technique are provided in
\cite{z10}.}

@@

<<slovenia>>=
x <- results$Slovenia
spval <- sctest(x$efp, functional = supLM(0.1))$p.value
@

To illustrate our method, we consider the case of Slovenia. After a period of
relative instability following the independence from Yugoslavia, Slovenia was
successfull in realigning its economy and introduced a number of reforms leading
to a stable growth in the past years. The good economic performance made it
possible for Slovenia to enter the ERM~II in June 2004 and later on to introduce
the euro in January 2007. The data collected contains the period from January
1996 onwards. In the first few years, Slovenia experienced very high inflation
rates, but was successfull in reducing inflation to a much lower level.

A glance at the time series depicted in Figure~\ref{fig:slovenia1} of the monthly inflation
rates reveals a decline in the mean around the middle of the observed time
period. The $efp(\cdot)$ starts to cross the border as early as 2001 and reaches
the peak somewhere between 2003 and 2004. The $p$~value of the sup$\mathit{LM}$ test is
\Sexpr{round(spval, 8)} and clearly indicates at least one break.  The number of
breaks is calculated by means of the LWZ criteria, which chooses one break in
August 2003. The goodness-of-fit test depicted in Figure~\ref{fig:slovenia2} confirms
the choice of the GL distribution. The mean inflation rate was cut to only a
third of its previous value, whereas variance increased slightly and the
distribution slightly skewed to the left, but peaking at a lower value than
previously.

In this year, Slovenia introduced a number of financial reforms
and severely decreased the year on year growth in the M3 monetary aggregate.
Combined, these actions were enough to ensure the successfull participation in
the ERM~II a year later and the introduction of the euro two years later.

@@

@@

\readme{Jesus: include p-values in table; no place left; Jesus: opening seems a
bit confusing and obscure}

Table~\ref{tab:results1} presents the break dates and the parameter values of
the segments as well as the dates of entry into the ERM/ERM~II and the date of
the euro introduction.

@@

\begin{table}[t!]
\begin{center}
\footnotesize
\begin{tabular}{llrrrlll}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline

@@

The first group that draws particular attention is a subgroup of the PIIGS
(Portugal, Ireland, Italy, Greece and Spain) countries consisting of Ireland and
Spain. Both show a break in the summer of 2008 following the financial crisis
and the severe problems in both countries concerning the real estate situation.
In both countries this had a strong deflationary impact, with Ireland actually
showing negative inflation rates for an extended period of time; in Spain, which
was affected by a housing bubble of roughly the same size as Ireland, the
housing prices did not adjust so abruptly and in such a degree. This
deflationary trend, which for a time affected all countries, can be seen in
Figure~\ref{fig:glmean}, which depicts a rolling mean estimation, where the mean
is calculated using the first moment of a fitted GL distribution. There is
evidence of a decline in mean inflation, which returns to its previous path
roughly a year later.

@@

Another interesting group consists -- with the exception of Austria and
Luxembourg -- of the countries of the former DM-zone. All these countries have a
very low inflation rate in both mean and variance and their inflation dynamics
did not change.

@@

In the core-zone, we see a decrease in the lower quartil from \Sexpr{yrs[1]} to \Sexpr{yrs[4]}, with no change in the median inflation rate, which is due to the fact, that there is no break after \Sexpr{yrs[2]} for these countries; however, the interquartile range decreased by almost 20\%. The variance increased slightly (but with a 15\% decrease in interquartile range), whereas skewness stayed symmetric in the median, with almost no change in the interquartile range.

@@

Our results were obtained using 
\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} \citep{R} with the packages
\pkg{glogis}~\Sexpr{packageDescription("glogis")["Version"]},
\pkg{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \citep{Zeileis+Leisch+Hornik:2002}, and
\pkg{fxregime}~\Sexpr{packageDescription("fxregime")["Version"]} \citep{z10},
all of which are available under the General Public License (GPL) from the
Comprehensive \proglang{R} Archive Network (\url{http://CRAN.R-project.org/}). Package \pkg{strucchange} provides
various techniques for inference in structural-change settings which are complemented
with additional maximum-likelihood-based methods by \pkg{fxregime}. Inference for the
generalized logistic model (that can be combined with \pkg{strucchange}/\pkg{fxregime})
is implemented in the new package \pkg{glogis} that also contains the data used, both
the raw monthly price indexes \citep[as provided by][]{oecd} and the seasonally adjusted
monthly inflation rates. The seasonal adjustment was performed in \pkg{X-12-ARIMA}~0.3
(\citealp{X-12-ARIMA}, see also \citealp{Findley+Monsell+Bell:1998}) through the interface
provided by \proglang{gretl}~1.9.5 (\citealp{gretl}, see also \citealp{Smith+Mixon:2006} for a review).

@@

\section{GL distribution}\label{sec:gld2}

The moments of the GL distribution as given in Equation~\ref{eq:dglogis} are
%
\begin{eqnarray}
\mathrm{E}(y) & = & \theta + \sigma (\gamma(\delta) - \gamma(1)) \\
\mathrm{Var}(y)  & = & \sigma^2(\gamma'(\delta)+\gamma'(1)) \\
\mathrm{Skew}(y) & = & \frac{\gamma''(\delta)-\gamma''(1)}{(\gamma'(\delta)+\gamma'(1))^{3/2}}
\end{eqnarray}
%
where $\gamma'(\cdot)$ and $\gamma''(\cdot)$ are the first and the second
derivatives of the digamma function, respectively. The log-likelihood is
%
\begin{eqnarray}
\log f(y|\theta,\sigma, \delta) & = & \log(\delta) - \log(\sigma) - \frac{1}{\sigma} (y-\theta) - (\delta+1) \nonumber \\
& & \times \log (1+\exp^{-\frac{y-\theta}{\sigma}}) \nonumber 
\end{eqnarray}
%
The resulting score function, where the scores are the derivatives of the
log-likelihood function, has three components $(s_{\theta}, s_{\sigma},
s_{\delta})$, with $\tilde{y} = \exp^{-\frac{y-\theta}{\sigma}}$. These are
employed in Equation~\ref{eq:efp} for the construction of the test statistics
and are a means of measuring the estimation error in a maximum likelihood
framework.

\begin{eqnarray}
s_{\theta}(y|\theta, \sigma, \delta) & =  &\frac{\delta \log f(y|\theta,\sigma, \delta)}{\delta \theta}   \\
&=&  \frac{1}{\sigma} - (\delta+1)\cdot\frac{\frac{1}{\sigma}\tilde{y}}{(1+\tilde{y})}  \nonumber \\
s_{\sigma}(y|\theta, \sigma, \delta) & = & \frac{\delta \log f(y|\theta,\sigma, \delta)}{\delta \sigma}   \\
& =&  -\frac{1}{\sigma} + \frac{1}{\sigma^2}(y-\theta) - (\delta+1) \times \frac{\frac{1}{\sigma^2}(y-\theta)\tilde{y}}{(1+\tilde{y})} \nonumber \\
s_{\delta}(y|\theta, \sigma, \delta) & = & \frac{\delta \log f(y|\theta,\sigma, \delta)}{\delta \delta}  \\
& =&   \frac{1}{\delta} - \log(1+\tilde{y}) \nonumber 
\end{eqnarray}
%
The scores of the parameters $\phi = (\theta, \sigma, \delta)$ sum up to 0: $\sum_{t=1}^n s(y_t|\hat{\phi}) ~=~ 0$. The first order condition is given by:
%
\begin{eqnarray}
\sum_{i=1}^n s(y_i|\hat{\theta}, \hat{\sigma}, \hat{\delta}) = 0
\end{eqnarray}

>>>>

However, a deeper inquiry into this problem is in need of a method able to
capture changes in inflation dynamics. We contribute to this debate by
developing a new method for testing for structural breaks in the dynamics of the
inflation rates and thereby upon the effect of the EMU on a number of European
countries both within and outside the eurozone. The framework developed in this
paper makes it possible to test for any significant changes of a countries
inflation dynamics and thus helps interpreting the effects of the EMU. We focus
on changes in the first three moments of these inflation rates, which we assume
to be distributed according to a generalized logistic distribution (GL). The GL
distribution is a suitable one due to its flexibility to capture nonsymmetry and
because of its fatter tails. For this purpose, we develop a new estimation
approach building on previous research by \citet{Zeileis+Hornik:2007}.

@@

Hence, a likelihood-based model is adopted that can also incorporate stylized
facts for return data, such as heavy tails and skewness. Here, the seasonally
adjusted returns $y_i$ ($i = 1, \dots, n$) are assumed to come from a
generalized logistic (GL) distribution \citep[see][]{johnson} with
three-dimensional parameter $\phi_i = (\theta_i, \sigma_i, \delta_i)^\top$ for
location~$\theta_i$, scale~$\sigma_i$, and shape~$\delta_i$ at time $i$. Structural change
techniques are then employed to check whether the parameter vector $\phi_i$
remains constant over time and, if that is not the case, when and how the
parameters change. This framework allows to trace breaks in the
evolution of mean, variance and skewness of the inflation series which may
possibly be linked to underlying regime changes, e.g., in monetary policy.

@@

In the following, details are provided about (1)~the GL distribution and its properties,
(2)~how it can be estimated (under parameter stability), (3)~how parameter stability
can be assesed, and (4)~how breakpoints can be estimated in the presence of parameter
instability.

@@

The logistic distribution is often used in econometrics in the
context of income and growth models due to its fatter tails compared to the 
normal distribution -- it has also been applied to the modeling of
inflation rates by \citet{batchelor}. Its generalization -- which additionally
allows for asymmetries -- has been applied to analyze risk \citep[e.g., by][]{tolikas}
but has not yet been applied -- to the best of our knowledge -- to inflation rates.
However, it will prove to be a rather simple model that fits HICP inflation rates
quite well.

The type~I GL distribution (also known as skew-logistic distribution), as defined in \citet{johnson},
has the log-density function
%
\begin{eqnarray}
%% f(y ~|~ \theta, \sigma, \delta) & = &
%% \frac{\frac{\delta}{\sigma} \cdot \exp\{-\frac{y - \theta}{\sigma}\}}{(1+\exp\{-\frac{y - \theta}{\sigma}\})^{(\delta+1)}} 
\ell(y ~|~ \theta, \sigma, \delta) & = & \log(\delta) - \log(\sigma) - \frac{y - \theta}{\sigma} -
                                         (\delta + 1) \cdot \log\left(1 + \exp\left\{-\frac{y - \theta}{\sigma}\right\}\right)
\label{eq:dglogis}
\end{eqnarray}
%
with location~$\theta$, scale~$\sigma > 0$ and shape~$\delta > 0$. For $\delta = 1$ the
distribution simplifies to the logistic distribution, for $\delta < 1$ or $> 1$ it is
left- or right-skewed, respectively. The corresponding first three moments are
%
\begin{eqnarray}
\mathrm{E}(y) & = & \theta + \sigma (\gamma(\delta) - \gamma(1)) \\
\mathrm{Var}(y)  & = & \sigma^2(\gamma'(\delta)+\gamma'(1)) \\
\mathrm{Skew}(y) & = & \frac{\gamma''(\delta)-\gamma''(1)}{(\gamma'(\delta)+\gamma'(1))^{3/2}}
\end{eqnarray}
%
where $\gamma(\cdot)$, $\gamma'(\cdot)$, and $\gamma''(\cdot)$ are the digamma function and its
first and second derivative, respectively.

The corresponding score function, i.e., the derivative of the log-density with respect
to the parameter vector, is given by
%
\begin{equation}
s(y ~|~ \phi) ~=~ \frac{\partial \ell(y ~|~ \phi)}{\partial \phi}  ~=~ \left(
  \begin{gathered}
    \frac{1}{\sigma} - \frac{(\delta + 1) \tilde{y}}{\sigma (1 + \tilde{y})} \\
    \left\{ \frac{1}{\sigma} - \frac{(\delta + 1) \tilde{y}}{\sigma (1 + \tilde{y})} \right\} \cdot \frac{y - \theta}{\sigma} -\frac{1}{\sigma} \\
    \frac{1}{\delta} - \log(1 + \tilde{y})
  \end{gathered} \right)
\end{equation}
%
where $\phi = (\theta, \sigma, \delta)^\top$ and
$\tilde{y} = \exp\{- (y - \theta)/\sigma\}$.

@@

Under the assumption that $y_i \sim \mathit{GL}(\theta, \sigma, \delta)$
for $i = 1, \dots, n$ independently -- i.e., are independent realizations from
a GL distribution with parameter vector $\phi = (\theta, \sigma, \delta)^\top$ --
the parameters can be estimated as usual by maximum likelihood (ML):
$\hat \phi ~=~ \argmax_\phi \sum_{i = 1}^n \ell(y_i | \phi)$. The corresponding
first order condition is $\sum_{i = 1}^n s(y_i | \hat \phi) = 0$.

To guard the inference against potential misspecification, e.g., autocorrelation or
misspecification of higher moments of the distribution, one can treat $\hat \phi$
as the quasi-maximum-likelihood (QML) estimator and adjust the inference by using
heteroskedasticity and autocorrelation consistent (HAC) covariance matrix
estimators \citep[see e.g.,][]{Andrews:1991}.

If the parameters are potentially varying over time --
$y_i \sim \mathit{GL}(\theta_i, \sigma_i, \delta_i)$ -- then $\hat \phi$ is the (Q)ML
estimator under the null hypothesis of parameter stability
%
\begin{equation} \label{eq:h0}
  H_0:~ \phi_i ~=~ \phi_0 \quad (i = 1, \dots, n)
\end{equation}
%
which should be tested against the alternative $H_1$ that at least one of the
parameters changes over time.

\subsection{Test for structural change}

The null hypothesis of parameter stability (\ref{eq:h0}) can be assessed using the empirical
scores $s(y_i | \hat \phi)$ as measures of model deviation \citep{Zeileis+Hornik:2007} from the
model fit under $H_0$. Systematic deviations over time from the full sample estimates $\hat \phi$
can then be captured by cumulative sums of the empirical scores in an empirical fluctuation
process $\mathit{efp}(t)$:
%
\begin{equation} \label{eq:efp}
\mathit{efp}(t) ~=~ \hat V^{-1/2} n^{-1/2} \sum_{i = 1}^{\lfloor nt \rfloor}
  s(y_i | \hat \phi) \quad (0 \leq t \leq 1),
\end{equation}
%
where $\hat V$ is some consistent estimator of the variance of the scores. Below,
we use the outer product of gradients $s(y_i | \hat \phi)$ to estimate $\hat V$.
Alternatively, HAC estimators could be used leading to the same qualitative
results for the data under investigation.

As usual in structural change analysis, a functional central limit theorem holds for
$\mathit{efp}(\cdot)$, which converges to a 3-dimensional Brownian
bridge: $\mathit{efp}(\cdot) ~\overset{d}{\rightarrow}~ W^0(\cdot)$. Based on
this various types of test statistics can be computed
(see \citealp{Zeileis:2005} for details). In the following, the 
sup$\mathit{LM}$ test of \cite{Andrews:1993} is employed, which performs particularly
well for single shift alternatives and if several of the three distribution
parameters change simultaneously. The test statistic with 10\% trimming is given by
%
\begin{eqnarray}
\sup_{t \in [0.1, 0.9]} \frac{\Vert \mathit{efp}(t) \Vert_2^2}{t(1-t)} \nonumber
\end{eqnarray}
%
and the appropriate $p$~values can be computed using from the corresponding
limiting distribution -- with $\mathit{efp}(t)$ replaced by $W^0(t)$.

@@

If the null hypothesis of parameter stability (\ref{eq:h0}) is rejected, a natural
strategy is to assume that there are $B$ breakpoints with stable parameters within
each of the resulting segments. \cite{bp03} have established a rigorous inference
framework in this situation for least squares estimation which has been extended by
\cite{Zeileis+Shah+Patnaik:2010} to (Q)ML estimation. Here, we follow the same ideas
and maximize the full segmented log-likelihood
%
\begin{equation*}
  \sum_{b = 1}^{B+1} \sum_{i = \tau_b + 1}^{\tau_b} \ell(y_i ~|~ \phi^{(b)})
\end{equation*}
%
for joint estimation of the breakpoints $\tau_1, \dots, \tau_B$ and the segment-specific
GL parameters $\phi^{(b)}$ ($b = 1, \dots, B + 1$). Following the recommendations
of \cite{bp03}, a modified Bayes information criterion \citep[LWZ, proposed by][]{liu}
is employed selecting the number of breakpoints $B$.

@@

<<slovenia>>=
x <- results$Slovenia
x$efp2 <- gefp(x$model0, fit = NULL, vcov = kernHAC)
spval <- c(sctest(x$efp, functional = supLM(0.1))$p.value,
  sctest(x$efp2, functional = supLM(0.1))$p.value)
spval <- ifelse(spval < 0.001, "$< 0.001$", format(round(spval, digits = 4), nsmall = 4))
@

To illustrate the proposed strategy, the case of Slovenia is considered. After a period of
relative instability following the independence from Yugoslavia, Slovenia was
successfull in realigning its economy and introduced a number of reforms leading
to a stable growth in recent years. The good economic performance made it
possible for Slovenia to enter the ERM~II in June~2004 and later on to introduce
the euro in January~2007. The seasonally adjusted inflation series for Slovenia encompasses
the time from \Sexpr{start(x$data)} to \Sexpr{end(x$data)} and is depicted in the left
panel of Figure~\ref{fig:slovenia1}.

@@

Assuming a single stable set of parmaters (\ref{eq:h0}) is clearly not valid, as the
sequence of $\mathit{LM}$~statistics (see right panel of Figure~\ref{fig:slovenia1}) clearly
exceeds its 5\% critical value (red horizontal line), leading to a highly significant
$p$-value \Sexpr{spval[1]}. If a quadratic spectral kernal HAC estimator \citep{Andrews:1991}
would have been used to adjust (\ref{eq:efp}), the resulting $p$-value would be somewhat
larger with \Sexpr{spval[2]} but still clearly significant.

As there is evidence for at least one break, the LWZ criterion is employed for $B = 1, 2, \dots$
breakpoints with a minimal segment size of two years per segment. It assumes its minimum for $B = 1$
associated with the breakpoint \Sexpr{breakdates(x$breakpoints, breaks = x$nbreaks)}, depcited
by a vertical dashed line in both panels of Figure~\ref{fig:slovenia1}. The resulting segmented
fitted mean (horizontal blue lines in the left panel Figure~\ref{fig:slovenia1}) shows that
in before the break Slovenia experienced very high inflation rates but was successful in reducing
inflation to a much lower level afterwards.

This decrease in inflation is also conveyed by Figure~\ref{fig:slovenia2} that shows the
observed histograms of inflation rates before and after the breakpoint along with the
fitted GL distribution. This highlights that along with the decrease in mean level, the variance increases
somewhat and -- probably more interestingly -- the distribution changes from being right-skewed to
slightly left-skewed. This confirms that very high inflation rates that occurred in several
months before the breakpoint could be avoided afterwards. Furthermore, Figure~\ref{fig:slovenia2}
shows that the GL distribution fits the observed data very well, both before and after the
breakpoint, with $p$~values from $\chi^2$ goodness-of-fit tests of
\Sexpr{format(round(summary(x$model[[1]])$chisq.test$p.value, digits = 3), nsmall = 3)} and
\Sexpr{format(round(summary(x$model[[2]])$chisq.test$p.value, digits = 3), nsmall = 3)}, respectively.

The estimated breakpoint can be matched very well with the timing of financial reforms
in Slovenia that severely decreased the annual growth in the M3 monetary aggregate.
These success of these actions ensured the Slovenian participation in
the ERM~II only a year later and the introduction of the euro two years later.

@@

The modeling strategy proposed in Section~\ref{sec:model} is applied to each of the
countries included in the data set introduced in Section~\ref{sec:data}.
Table~\ref{tab:results1} reports the estimated breakpoints along with the segment-specific
moments (mean, variance, skewness) implied by the corresponding parameter
estimates. Furthermore, entry dates for ERM/ERM~II and the euro are included as
additional information. For ease of interpretation, countries with similar patterns of
change are grouped together in Table~\ref{tab:results1} (and ordered alphabetically
within each group).

If there is only a single segment, the sup$\mathit{LM}$ test was non-significant at 5\%
level. In significant cases, the number of breakpoints was selected via the LWZ criterion
and subsequently all parameters (breakpoints and segment-specific GL parameters) were
estimated by ML. No HAC~correction was employed as there was only negligible autocorrelation
in the empirical scores after inclusion of the breakpoints (if any).

@@

\begin{table}[p!]
\begin{center}
\footnotesize
\begin{tabular}{llrrrlll}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline

@@

\fixme{Z: I have reordered the groups in Table~1 which tries to be roughly in the order
of joining EU/ERM/EMU. Probably the following interpretations should start with a
very short overview of the groupings.}

The first group that draws particular attention is a subgroup of the PIIGS countries
(Portugal, Ireland, Italy, Greece, Spain) consisting of Ireland and
Spain. Both show a break in the summer of 2008 following the financial crisis
and the severe problems in both countries concerning the real estate situation.
In both countries this had a strong deflationary impact, with Ireland actually
showing negative inflation rates for an extended period of time; in Spain, which
was affected by a housing bubble of roughly the same size as Ireland, the
housing prices did not adjust so abruptly and in such a degree. This
deflationary trend, which for a time affected all countries, can be seen in
Figure~\ref{fig:glmean}, which depicts a rolling mean estimation, where the mean
is calculated using the first moment of a fitted GL distribution. There is
evidence of a decline in mean inflation, which returns to its previous path
roughly a year later.

@@

Another interesting group consists -- with the exception of Austria and
Luxembourg -- of the countries of the former DEM-zone. All these countries have a
very low inflation rate in both mean and variance and their inflation dynamics
did not change.

@@

In the core-zone, we see a decrease in the lower quartile from \Sexpr{yrs[1]} to
\Sexpr{yrs[4]}, with no change in the median inflation rate, which is due to the
fact, that there is no break after \Sexpr{yrs[2]} for these countries; however,
the interquartile range decreased by almost 20\%. The variance increased
slightly (but with a 15\% decrease in interquartile range), whereas skewness
stayed symmetric in the median, with almost no change in the interquartile
range.

@@

Our results were obtained using 
\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} \citep{R} with the packages
\pkg{glogis}~\Sexpr{packageDescription("glogis")["Version"]},
\pkg{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \citep{Zeileis+Leisch+Hornik:2002}, and
\pkg{fxregime}~\Sexpr{packageDescription("fxregime")["Version"]} \citep{Zeileis+Shah+Patnaik:2010},
all of which are available under the General Public License (GPL) from the
Comprehensive \proglang{R} Archive Network (\url{http://CRAN.R-project.org/}). Package \pkg{strucchange} provides
various techniques for inference in structural-change settings which are complemented
with additional maximum-likelihood-based methods by \pkg{fxregime}. Inference for the
generalized logistic model (that can be combined with \pkg{strucchange}/\pkg{fxregime})
is implemented in the new package \pkg{glogis} that also contains the data used, both
the raw monthly price indexes \citep[as provided by][]{oecd} and the seasonally adjusted
monthly inflation rates. The seasonal adjustment was performed in \pkg{X-12-ARIMA}~0.3
(\citealp{X-12-ARIMA}, see also \citealp{Findley+Monsell+Bell:1998}) through the interface
provided by \proglang{gretl}~1.9.5 (\citealp{gretl}, see also \citealp{Smith+Mixon:2006} for a review).

@@
